master thesis

### 2018-03-22 Paper - Latex/Paper - Skizze - v4.tex
Paper - Introduction - Anmerkungen von Thilo eingearbeitet

  We will test the existing maximum parsimony algorithms Fitch \cite{Fitch1971} and Sankoff 
    \cite{Sankoff1975} for this task and estimate their predictive power. The present tree of OTL is 
    not binary. A tree is binary (or bifurcated), when every node has at most two children otherwise 
    it is multifurcated \cite{Felsenstein2003}. \\
  Maximum parsimony methods are developed for phylogenies, which are usually depicted as binary trees.
    Parsimonious in phylogeny refers to favoring the tree that needs the least evolutionary change 
    to explain the observed data. In our case, it is about the change of states 'is free-living' or 
    'is parasitic'. \\
    % Parsimonious in this context means to minimize the number of transitions between species states 
    % (free-living or parasitic). 
  The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
    package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
    method and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
  This achieves that we can predict ancestral states and unknown states of living species for large 
    non-binary relatives trees. \\

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
    \cite{Camin1965}. Different methods have been developed and the question is which method is the 
    most suitable for our problem. \\
  \anmerkungstext{hier mehr dazu? Likelihood vs parsiomony... das befindet sich momentan im Methoden 
    Teil: Ancestral state reconstruction methods} \\
  % The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\

@@

  To accomplish this task, a large phylogenetic tree and information about the current species states 
    is needed. \\
  The biggest 'phylogenetic tree' is a synthesis of phylogenetic trees filled with a taxonomic trees 
    given by Open Tree of Life (OTL) \cite{Hinchliff2015}. %This tree is not binary and therefore the 
    % developed algorithms are not directly applicable. Therefore, an algorithm must be found that can 
    % work with non-binary trees. \\
  For the information about the current states of the species we use the interaction database Global 
    Biotic Interactions (GloBI) \cite{Poelen2014}. The data in GloBI are stored as interactions. 
    Useful for us e.g. Species A parasitize species B. From this we conclude that species A is 
    parasitic and species B free-living. In the case parasite parasitizes parasite, we get conflicting 
    states for the second species. We solve this by preferring parasitic. \\

  At this point a few words to the term parasitic. There are different definitions. Since we use Globi 
    to classify species, we use their definition of parasitism. Again, in GloBi, ontobee definitions 
    are used.
  The interaction has parasite is defined as: "An interaction relationship between two organisms 
    living together in more or less intimate association in a relationship in which association is 
    disadvantageous or destructive to one of the organisms." \cite{...}. This definition includes: 
    ecto- and endoparasites, parasitoids, kleptoparasites and pathogenes.
  \todotext{Mungall, C., (2017). Definition for the interaction-term: "parasitised by; has parasite" 
  ob ontobee.org. Last checked: 24.07.2017 at 
  \hyperlink{ontobee.org/ontology/RO?iri=http://purl.obolibrary.org/obo/RO_0002445}
  {ontobee.org/ontology/...}} \\

  This paper thus consists of the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Perform reconstructing on the Eucaryotic synthesis tree of OTL. \\
  Point 1 consists of a simulation of our data situation and the application of various possible 
    algorithms to these. To evaluate them based on the results. \\
  The one by Louca et al. implemented Sankoff algorithm is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states. 
    And then an evaluation of the results.

@@

      This database consists of entries of the form: species A (source) interacts with B (target). \\
      A number of interactions have been identified\footnote{\hyperlink{
        https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
        }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}}
        ,including those that the species source or target has become a parasite or a free-living 
        species from the biological perspective. These are the following:
      \begin{itemize}
        \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
          hasParasite, hostOf
        \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
        \item parasite source: parasiteOf, pathogenOf
        \item parasite target: hasParasite, hasPathogen
      \end{itemize}
      Two lists are formed: parasites and free-livings, and the source or targets of an interaction
        are added to these.

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & XXXXX \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}
      It also follows from this table that the most complex model is the best. In general, the BIC 
        values are smaller than those of the multifurcation models. The modeling here is thus better. \\
      Again, the calculation of finer models (eg family) was too expensive. \\

@@

    As presented, we compare two methods in our simulation to their prediction accuracy: Fitch and 
      Sankoff. \\
    We examine different parameters. In figure \ref{fig:influence of unknown data} is an overview of 
      the results. \\
    The first column describes the distributions of free-livings and parasites with a given threshold 
      for the respective simulations to the right. \\
    The middle column investigates the influence of the unknown states, the right the influence of the 
      strength of the multifurcation. \\
    The y-axes indicate the percentage of correctly predicted states (including known states). On the 
      x-axis the percentage of forgotten states or missing internal nodes. \\
    Each point corresponds to the average of one hundred simulations, each with 10,000 leaf nodes. \\
    For the middle column we set the strength of the multifurcations to 0.95\% similar to the real 
      data and in the right column the amount of the unknowns to 0.95\% also similar to the real data. \\

    As \colorbox{red}{you} can see both algortithms are always over 50\% and therefore better than guessing. Moreover, 
      they are usually close to each other, with Sankoff always makes better predictions except for 
      equally distributed states as Fitch. \\
    \colorbox{red}{You} can see that the distribution of states has a strong influence on the prediction. The more 
      evenly distributed, the harder it is to predict. \\
    If more than 60\% of internal nodes are missing, Fitch breaks significantly in his prediction 
      compared to Sankoff.

@@

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            correct predicted $\sim$ taxa & XXXXX & XXXXX & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa + depth & 117703 & XXXXX & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa * depth & 117592 & XXXXX & XXXXX & \cellcolor{green!60}XXXXX \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{Residuals of cross validation prediction}
        \label{table:Residuals cross validation} 
      \end{table}

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            correct predicted $\sim$ taxa & 117936 & 112242 & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa + depth & 117776 & XXXXX & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa * depth & XXXXX & XXXXX & XXXXX & \cellcolor{green!60}XXXXX \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of cross validation prediction}
        \label{table:BIC cross validation} 
      \end{table}

      Resiudals:
      \begin{itemize}
        \item correct predicted $\sim$ 1: 120325
        \item correct predicted $\sim$ kingdom: 117877
        \item correct predicted $\sim$ phylum: 111466
      \end{itemize}

>>>>

  We will test the existing maximum parsimony algorithms Fitch \cite{Fitch1971} and Sankoff 
    \cite{Sankoff1975} for this task and estimate their predictive power. The present tree structure 
    of OTL is not binary. A tree is binary (or bifurcated), when each node has at most two children 
    otherwise it is multifurcated \cite{Felsenstein2003}. \\
  % \anmerkungstext{auch hier kannst du multiifurcated direkt erklären: that is it each not has 
    % multiple (n >= 3) children (Thilo)}
  Maximum parsimony methods are developed for phylogenies, which are usually depicted as binary trees.
    Parsimonious in phylogeny refers to favoring the tree that needs the least evolutionary change 
    to explain the observed data. In our case, it is about the change of states 'is free-living' or 
    'is parasitic'. \\
    % Parsimonious in this context means to minimize the number of transitions between species states 
    % (free-living or parasitic). 
  The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
    package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
    method and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
  This achieves that we can predict ancestral states and unknown states of living species for large 
    non-binary relatives trees. \\

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
    \cite{Camin1965}. Different methods have been developed and the question is which method is the 
    most suitable for the problem at hand: The ancestral state reconstruction for a huge 
    multifurcated tree with \colorbox{red}{binary/two discrete} states. \\
  \anmerkungstext{hier mehr dazu? Likelihood vs parsiomony... das befindet sich momentan im Methoden 
    Teil: Ancestral state reconstruction methods} \\
  % Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
  %   The first is to explain the current state with the least number of state changes between an 
  %     ancestor and his child, this is called parsimonious. \\
  %   The other class she presents involves modeling the character evolution as a stochastic process and 
  %     using the likelihoods to compute the possible ancestral character states. This is generally done 
  %     with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

  % The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\

@@

  To accomplish this task, a large phylogenetic tree and information about the current species states 
    is needed. \\
  The biggest 'phylogenetic tree' is a synthesis of phylogenetic trees filled with a taxonomic trees 
    given by Open Tree of Life (OTL) \cite{Hinchliff2015}. %This tree is not binary and therefore the 
    % developed algorithms are not directly applicable. Therefore, an algorithm must be found that can 
    % work with non-binary trees. \\
  For the information about the current states of the species we use the interaction database Global 
    Biotic Interactions (GloBI) \cite{Poelen2014}. The data in GloBI are stored as interactions e.g. 
    species A parasitize species B. From this we conclude that species A is parasitic and species B 
    free-living. \\

  At this point a few words to the term parasitic. There are different definitions. Since we use 
    GloBI to classify species, we use their definition of parasitism. Again, in GloBi, Ontobee 
    definitions are used \cite{Xiang2011}.
  The interaction \textit{has parasite} is defined as: "An interaction relationship between two 
    organisms living together in more or less intimate association in a relationship in which 
    association is disadvantageous or destructive to one of the organisms." \cite{...}. This 
    definition includes: ecto- and endoparasites, parasitoids, kleptoparasites and pathogenes.
  \todotext{Mungall, C., (2017). Definition for the interaction-term: "parasitised by; has parasite" 
  ob ontobee.org. Last checked: 24.07.2017 at 
  \hyperlink{ontobee.org/ontology/RO?iri=http://purl.obolibrary.org/obo/RO_0002445}
  {ontobee.org/ontology/...}} \\

  The objectives of this work are the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Perform reconstructing on the Eucaryotic synthesis tree of OTL. \\
  The goal of Point 1 is to evaluate the possible methods based on a simulation of our data 
    situation. \\
  The one by Louca et al. implemented Sankoff algorithm is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states. 
    And then perform an evaluation of the results.

@@

      This database consists of entries of the form: species A (source) interacts with B (target). \\
      A number of interactions have been identified\footnote{\hyperlink{
        https://github.com/jhpoelen/eol-globi-data/blob/master/eol-globi-lib/src/main/java/org/eol/globi/domain/InteractType.java
        }{https://github.com/jhpoelen/eol-globi-data/.../InteractType.java}}
        ,including those that the species source or target has become a parasite or a free-living 
        species from the biological perspective. These are the following:
      \begin{itemize}
        \item free-living source: preysOn, eats, flowersVisitedBy, hasPathogen, pollinatedBy, 
          hasParasite, hostOf
        \item free-living target: preyedUponBy, parasiteOf, visitsFlowersOf, pathogenOf, hasHost
        \item parasite source: parasiteOf, pathogenOf
        \item parasite target: hasParasite, hasPathogen
      \end{itemize}
      Of these interactions, e.g. species A parasitize species B, the state of the species is 
        determined, here is species A parasitic and species B free-living. The case a parasite 
        conquers (parasitizes) another parasite yields conflicting states for the second species. 
        This is solved by preferring parasitic. \\
      Two lists are formed: parasites and free-livings, and the source or targets of an interaction
        are added to these.

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & \cellcolor{green!45}484681 \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}
      It also follows from this table that the most complex model is the best. In general, the BIC 
        values are smaller than those of the multifurcation models. The modeling here is thus better. \\
      Again, the calculation of finer models (eg order or family) was too expensive. \\

@@

    As presented, we compare two methods in our simulation to their prediction accuracy: Fitch and 
      Sankoff. \\
    We examine different parameters. In figure \ref{fig:influence of unknown data} is an overview of 
      the results. \\
    The first column describes the distributions of free-livings and parasites with a given threshold 
      for the respective simulations to the right. \\
    The middle column investigates the influence of the unknown states, the right the influence of the 
      strength of the multifurcation. \\
    The y-axes indicate the percentage of correctly predicted states (including known states). On the 
      x-axis the percentage of forgotten states or missing internal nodes. \\
    Each point corresponds to the average of one hundred simulations, each with 10,000 leaf nodes. \\
    For the middle column we set the strength of the multifurcations to 0.95~\% similar to the real 
      data and in the right column the amount of the unknowns to 0.95~\% also similar to the real data. \\

    As \colorbox{red}{you} can see both algortithms are always over 50~\% and therefore better than guessing. Moreover, 
      they are usually close to each other, with Sankoff always makes better predictions except for 
      equally distributed states as Fitch. \\
    \colorbox{red}{You} can see that the distribution of states has a strong influence on the prediction. The more 
      evenly distributed, the harder it is to predict. \\
    If more than 60~\% of internal nodes are missing, Fitch breaks significantly in his prediction 
      compared to Sankoff.

@@

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            correct predicted $\sim$ taxa & 117877 & 111466 & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa + depth & 117703 & XXXXX & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa * depth & 117592 & XXXXX & XXXXX & \cellcolor{green!60}XXXXX \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{Residuals of cross validation prediction}
        \label{table:Residuals cross validation} 
      \end{table}

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            correct predicted $\sim$ taxa & 117936 & 112242 & 111733 & XXXXX \\
            \hline
            correct predicted $\sim$ taxa + depth & 117776 & XXXXX & XXXXX & XXXXX \\
            \hline
            correct predicted $\sim$ taxa * depth & XXXXX & XXXXX & XXXXX & \cellcolor{green!60}XXXXX \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of cross validation prediction}
        \label{table:BIC cross validation} 
      \end{table}
